Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-07 submission.tex
captions

In this paper, we optimize a PoS tag set for the task of data-driven dependency
parsing of Norwegian. We report on a set of experiments where various
morphological distinctions are introduced to the PoS annotations and evaluated
both intrinsically, i.e., in terms of PoS tagging accuracy, and extrinsically,
in terms of parsing accuracy. Our results show that the introduction of
morphological distinctions not present in the original tag set, whilst
compromising tagger accuracy, actually leads to significantly improved parsing
accuracy. This optimization allows us to bypass the additional step of
morphological analysis, framing the whole preprocessing problem as a simple
tagging task.

@@

\paragraph{Predicted vs gold tags}
As we seek to quantify the effects of PoS tagging in a realistic setting, i.e.,
in application to raw text, we choose to evaluate the parser on automatically
assigned PoS tags. For training, however, we have two options: using either
gold standard tags or automatically predicted tags. In order to settle on a
configuration, we conducted experiments with training on both gold tags 
and predicted tags to see how the performance differs. The
results of these experiments, shown in Table \ref{tagconfigeval}, reveal that
the combination of training and testing on predicted tags is clearly superior
to training on gold standard tags and testing on predicted tags. Consequently,
the parser is both trained and tested on automatically predicted tags in the
remainder of our experiments.\footnote{Though some parsers also make use of
  morphological features, we removed all morphological features beyond the PoS
  tags in order to simulate a realistic setting.}

@@

The modified tag sets used in the experiments reported in this section are
defined as combinations of PoS tags and morphological features already
available in the gold annotations of the treebank. In effect, we redefine the
tags comprising the gold standard as provided by NDT. The best performing
configuration can be seens as tag set specifically optimized for dependency
parsing. Note that the tag selection process itself can be seen as
semi-automatic; while we for each introduced distinction empirically evaluate
its impact on parsing performance -- as to see whether it is worth including in
the final configuration -- the process is manually guided by linguistic
considerations regarding which combinations to evaluate in the first place.
Although our expressed goal is to identify a tag set optimized for the
downstream task of dependency parsing, we will only consider tag set
modifications we deem linguistically sensible.

By introducing more fine-grained linguistically motivated distinctions in the tag
set, we increase the linguistic information represented in the tags, which may
assist the parser in recognizing and generalizing syntactic patterns.
However, this would also increase the complexity of the tagging task, 
which can be expected to lead to a drop in tagger accuracy.
However, the best tagging, evaluated in isolation, does not necessarily lead to the
best parse, hence it is interesting to investigate how tag set modifications
affect this interplay.

@@

\subsection{Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its full set of morphological features, thereby mapping the original tag set to a
new maximally fine-grained tag set, given the annotations available in the
treebank. This resulted in a total of 368 tags,
hereafter referred to as the \emph{full} tag set.
The two initial tag sets, i.e., the original tag set comprising 19 tags and the
full tag set comprising 368 tags, thus represent two extremes in terms of
granularity.
To establish some initial points of reference for how tag set granularity affects the
performance of tagging and parsing on NDT, we trained and tested a full pipeline with both of
these intital tags sets. The results are reported in Table \ref{inittagseteval}.
Unsurprisingly, we see that the tagger accuracy
plummets when we move from the original to the full tag set. While the MFT
tagging baseline for the original tag set is 94.14\%, it drops by almost 9
percentage points to 85.15\% for the full tag set. TnT reports an accuracy of
97.47\% on the original tag set, which is reduced to 93.48\% for the full tag
set. These results confirm our hypothesis that the very high level linguistic
information
in the full, fine-grained tag set comes at the expense of reduced tagger
performance.  In spite of this, however, we see that the additional information
in the full tag set still improves the parser performance. With the original tag set, Mate
yields 87.01\% LAS and 90.19\% UAS, which increases to 87.15\% and
90.39\%, respectively, using the full tag set. As we are looking for
linguistically informed distinctions that improve the syntactic parsing, these
results are promising and indicate that additional linguistic
information assists syntactic parsing, motivating the optimization of the
existing PoS tag set.

@@

\begin{table}
    \centering
    \smaller[1]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature(s)} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS}
        \\
        \midrule
        --- & 97.47\% & 87.01\% & 90.19\% \\
        Case & \textbf{97.48\%} & 87.63\% & 90.72\% \\
        Definiteness & 97.00\% & 88.27\% & 91.42\% \\
        Gender & 96.09\% & 87.21\% & 90.36\% \\
        Number & 96.37\% & 87.97\% & 91.00\% \\
        Type & 96.92\% & 88.07\% & 91.11\% \\
        Case \& definiteness & 97.03\% & 88.39\% & 91.44\% \\
        Type \& case & 96.92\% & 88.46\% & 91.51\% \\
        Type \& definiteness & 96.99\% & 88.44\% & 91.48\% \\
        %Type \& number & 96.37\% & 87.95\% & 90.95\% \\
        Type, case \& definiteness & 97.05\% & \textbf{88.81\%} &
        \textbf{91.73\%} \\
        %Type, definiteness \& number & 96.46\% & 88.06\% & 91.12\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of experiments with modified PoS tags for nouns.}
    \label{substresults}
\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature(s)} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        --- & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        Mood & 97.43\% & 87.04\% & 90.19\% \\
        Tense & 97.30\% & 86.97\% & 90.18\% \\
        Voice & 97.45\% & 86.96\% & 90.09\% \\
        Mood \& tense & 97.31\% & 87.12\% & 90.31\% \\
        Voice \& tense & 97.28\% & 86.99\% & 90.15\% \\
        Mood, tense \& voice & 97.27\% & 86.83\% & 90.05\% \\
        Finiteness & 97.35\% & \textbf{87.30\%} & \textbf{90.43\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of experiments with modified PoS tags for verbs.}
    \label{verbresults}
\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature(s)} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS}
        \\
        \midrule
        --- & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        Definiteness & 96.84\% & 87.14\% & 90.29\% \\
        Degree & 97.41\% & \textbf{87.29\%} & \textbf{90.44\%} \\
        Gender & 96.89\% & 87.10\% & 90.25\% \\
        Number & 96.71\% & 86.99\% & 90.10\% \\
        Type & 97.40\% & 87.11\% & 90.25\% \\
        Definiteness \& degree & 96.81\% & 87.23\% & 90.39\% \\
        Definiteness \& gender & 96.31\% & 87.18\% & 90.39\% \\
        Definiteness \& number & 96.78\% & 87.27\% & \textbf{90.44\%} \\
        %Degree \& gender & 96.87\% & 87.00\% & 90.16\% \\
        %Degree \& number & 96.76\% & 87.13\% & 90.26\% \\
        %Definiteness, degree \& number & 96.81\% & 87.14\% & 90.30\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of experiments with modified PoS tags for adjectives.}
    \label{adjresults}
\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        --- & 97.47\% & 87.01\% & 90.19\% \\
        Definiteness & 97.49\% & \textbf{87.30\%} & \textbf{90.42\%} \\
        Gender & 97.28\% & 87.09\% & 90.31\% \\
        Number & 97.49\% & 87.04\% & 90.18\% \\
        Type & \textbf{97.61\%} & 87.00\% & 90.11\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of experiments with modified PoS tags for determiners.}
    \label{detresults}
\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature(s)} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        --- & 97.47\% & 87.01\% & 90.19\% \\
        Case & 97.50\% & 87.08\% & 90.21\% \\
        Gender & 97.48\% & 87.06\% & 90.23\% \\
        Number & 97.49\% & 87.21\% & 90.33\% \\
        Person & 97.49\% & 87.22\% & 90.32\% \\
        Type & 97.48\% & 87.19\% & 90.40\% \\
        Number \& person & 97.49\% & 96.98\% & 90.16\% \\
        Type \& case & 97.51\% & \textbf{87.30\%} & \textbf{90.41\%} \\
        Type \& number & 97.49\% & 87.27\% & \textbf{90.41\%} \\
        Type \& person & 97.49\% & 87.00\% & 90.14\% \\
        Type, case \& number & \textbf{97.52\%} & 87.11\% & 90.36\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of experiments with modified PoS tags for pronouns.}
    \label{pronresults}
\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llllll@{}}
        \toprule
        \textbf{Data} & \textbf{Tag Set} & \textbf{MFT} & \textbf{Acc} &
        \textbf{LAS} & \textbf{UAS} \\
        \midrule
        \multirow{2}{*}{Dev}
        & Original & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        %Full & 85.15\% & 93.48\% & 87.15\% & 90.39\% \\
        & Optimized & 85.15\% & 96.85\% & \textbf{88.87\%} & \textbf{91.78\%} \\
        \midrule
        \multirow{2}{*}{Test}
        & Original & \textbf{94.22\%} & \textbf{97.30\%} & 86.64\% & 90.07\% \\
        %& Full & 84.27\% & 92.79\% & 87.03\% & 90.16\% \\
        & Optimized & 88.08\% & 96.35\% & \textbf{88.55\%} & \textbf{91.41\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the optimized tag set,
        compared to the original coarse tag set. Trained and tested using
        automatically assigned tags from TnT.}
    \label{finalresults}
\end{table}

>>>>

In this paper, we optimize a PoS tag set for the task of data-driven dependency
parsing of Norwegian. We report on a set of experiments where various
morphological distinctions are introduced to the PoS annotations and evaluated
both intrinsically, i.e., in terms of PoS tagging accuracy, and extrinsically,
in terms of parsing accuracy. Our results show that the introduction of
morphological distinctions not present in the original tag set, whilst
compromising tagger accuracy, actually leads to significantly improved parsing
accuracy. This optimization also allows us to bypass the additional step of
morphological analysis, framing the whole preprocessing problem as a simple
tagging task.

@@

\paragraph{Predicted vs gold tags}
As we seek to quantify the effects of PoS tagging in a realistic setting, i.e.,
in application to raw text, we choose to evaluate the parser on automatically
predicted PoS tags. For training, however, we have two options: using either
gold standard tags or automatically predicted tags. In order to settle on a
configuration, we conducted experiments with training on both gold tags 
and predicted tags to see how the performance differs. The
results of these experiments, shown in Table \ref{tagconfigeval}, reveal that
the combination of training and testing on predicted tags is clearly superior
to training on gold standard tags and testing on predicted tags. Consequently,
the parser is both trained and tested on automatically predicted tags in the
remainder of our experiments.\footnote{Though some parsers also make use of
  morphological features, we removed all morphological features beyond the PoS
  tags in order to simulate a realistic setting.}

@@

The modified tag sets used in the experiments reported in this section are
defined as combinations of PoS tags and morphological features already
available in the gold annotations of the treebank. In effect, we redefine the
tags comprising the gold standard as provided by NDT. The best performing
configuration can be seen as tag set specifically optimized for dependency
parsing. Note that the tag selection process itself can be seen as
semi-automatic; while we for each introduced distinction empirically evaluate
its impact on parsing performance -- as to see whether it is worth including in
the final configuration -- the process is manually guided by linguistic
considerations regarding which combinations to evaluate in the first place.
Although our expressed goal is to identify a tag set optimized for the
downstream task of dependency parsing, we will only consider tag set
modifications we deem linguistically sensible.

The hope is that the introduction of more fine-grained linguistically motivated
distinctions in the tag set may assist the parser in recognizing and
generalizing syntactic patterns.  However, this would also increase the
complexity of the tagging task, which can be expected to lead to a drop in
tagger accuracy.  However, the best tagging, evaluated in isolation, does not
necessarily lead to the best parse, hence the aim of this section is to
investigate how tag set modifications affect this interplay.

@@

\subsection{Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its full set of morphological features, thereby mapping the original tag set to a
new maximally fine-grained tag set, given the annotations available in the
treebank. This resulted in a total of 368 tags,
hereafter referred to as the \emph{full} tag set.
The two initial tag sets, i.e., the original tag set comprising 19 tags and the
full tag set comprising 368 tags, thus represent two extremes in terms of
granularity.
To establish some initial points of reference for how tag set granularity affects the
performance of tagging and parsing on NDT, we trained and tested a full pipeline with both of
these intital tags sets. The results are reported in Table \ref{inittagseteval}.
Unsurprisingly, we see that the tagger accuracy
plummets when we move from the original to the full tag set. While the MFT
tagging baseline for the original tag set is 94.14\%, it drops by almost 9
percentage points to 85.15\% for the full tag set. TnT reports an accuracy of
97.47\% on the original tag set, which is reduced to 93.48\% for the full tag
set. These results confirm our hypothesis that the very high level linguistic
information
in the full, fine-grained tag set comes at the expense of reduced tagger
performance.  In spite of this, however, we see that the additional information
in the full tag set still improves the parser performance. With the original tag set, Mate
yields 87.01\% LAS and 90.19\% UAS, which increases to 87.15\% and
90.39\%, respectively, using the full tag set. These
preliminary results are promising in that thay indicate that additional linguistic
information assists syntactic parser, motivating further optimization of the
existing PoS tag set.

@@

\begin{table}
    \centering
    \smaller[1]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature(s)} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS}
        \\
        \midrule
        --- & 97.47\% & 87.01\% & 90.19\% \\
        Case & \textbf{97.48\%} & 87.63\% & 90.72\% \\
        Definiteness & 97.00\% & 88.27\% & 91.42\% \\
        Gender & 96.09\% & 87.21\% & 90.36\% \\
        Number & 96.37\% & 87.97\% & 91.00\% \\
        Type & 96.92\% & 88.07\% & 91.11\% \\
        Case \& definiteness & 97.03\% & 88.39\% & 91.44\% \\
        Type \& case & 96.92\% & 88.46\% & 91.51\% \\
        Type \& definiteness & 96.99\% & 88.44\% & 91.48\% \\
        %Type \& number & 96.37\% & 87.95\% & 90.95\% \\
        Type, case \& definiteness & 97.05\% & \textbf{88.81\%} &
        \textbf{91.73\%} \\
        %Type, definiteness \& number & 96.46\% & 88.06\% & 91.12\% \\
        \bottomrule
    \end{tabular}
    \caption{Tagging and parsing with modified PoS tags for nouns. The first
      row corresponds to using the original tag set unmodified.}
    \label{substresults}
\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature(s)} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        --- & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        Mood & 97.43\% & 87.04\% & 90.19\% \\
        Tense & 97.30\% & 86.97\% & 90.18\% \\
        Voice & 97.45\% & 86.96\% & 90.09\% \\
        Mood \& tense & 97.31\% & 87.12\% & 90.31\% \\
        Voice \& tense & 97.28\% & 86.99\% & 90.15\% \\
        Mood, tense \& voice & 97.27\% & 86.83\% & 90.05\% \\
        Finiteness & 97.35\% & \textbf{87.30\%} & \textbf{90.43\%} \\
        \bottomrule
    \end{tabular}
    \caption{Tagging and parsing with modified PoS tags for verbs.}
    \label{verbresults}
\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature(s)} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS}
        \\
        \midrule
        --- & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        Definiteness & 96.84\% & 87.14\% & 90.29\% \\
        Degree & 97.41\% & \textbf{87.29\%} & \textbf{90.44\%} \\
        Gender & 96.89\% & 87.10\% & 90.25\% \\
        Number & 96.71\% & 86.99\% & 90.10\% \\
        Type & 97.40\% & 87.11\% & 90.25\% \\
        Definiteness \& degree & 96.81\% & 87.23\% & 90.39\% \\
        Definiteness \& gender & 96.31\% & 87.18\% & 90.39\% \\
        Definiteness \& number & 96.78\% & 87.27\% & \textbf{90.44\%} \\
        %Degree \& gender & 96.87\% & 87.00\% & 90.16\% \\
        %Degree \& number & 96.76\% & 87.13\% & 90.26\% \\
        %Definiteness, degree \& number & 96.81\% & 87.14\% & 90.30\% \\
        \bottomrule
    \end{tabular}
    \caption{Tagging and parsing with modified PoS tags for adjectives.}
    \label{adjresults}
\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        --- & 97.47\% & 87.01\% & 90.19\% \\
        Definiteness & 97.49\% & \textbf{87.30\%} & \textbf{90.42\%} \\
        Gender & 97.28\% & 87.09\% & 90.31\% \\
        Number & 97.49\% & 87.04\% & 90.18\% \\
        Type & \textbf{97.61\%} & 87.00\% & 90.11\% \\
        \bottomrule
    \end{tabular}
    \caption{Tagging and parsing with modified PoS tags for determiners.}
    \label{detresults}
\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature(s)} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        --- & 97.47\% & 87.01\% & 90.19\% \\
        Case & 97.50\% & 87.08\% & 90.21\% \\
        Gender & 97.48\% & 87.06\% & 90.23\% \\
        Number & 97.49\% & 87.21\% & 90.33\% \\
        Person & 97.49\% & 87.22\% & 90.32\% \\
        Type & 97.48\% & 87.19\% & 90.40\% \\
        Number \& person & 97.49\% & 96.98\% & 90.16\% \\
        Type \& case & 97.51\% & \textbf{87.30\%} & \textbf{90.41\%} \\
        Type \& number & 97.49\% & 87.27\% & \textbf{90.41\%} \\
        Type \& person & 97.49\% & 87.00\% & 90.14\% \\
        Type, case \& number & \textbf{97.52\%} & 87.11\% & 90.36\% \\
        \bottomrule
    \end{tabular}
    \caption{Tagging and parsing with modified PoS tags for pronouns.}
    \label{pronresults}
\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llllll@{}}
        \toprule
        \textbf{Data} & \textbf{Tag Set} & \textbf{MFT} & \textbf{Acc} &
        \textbf{LAS} & \textbf{UAS} \\
        \midrule
        \multirow{2}{*}{Dev}
        & Original & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        %Full & 85.15\% & 93.48\% & 87.15\% & 90.39\% \\
        & Optimized & 85.15\% & 96.85\% & \textbf{88.87\%} & \textbf{91.78\%} \\
        \midrule
        \multirow{2}{*}{Test}
        & Original & \textbf{94.22\%} & \textbf{97.30\%} & 86.64\% & 90.07\% \\
        %& Full & 84.27\% & 92.79\% & 87.03\% & 90.16\% \\
        & Optimized & 88.08\% & 96.35\% & \textbf{88.55\%} & \textbf{91.41\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the optimized tag set,
        compared to the original NDT coarse tag set. The parser is both trained and tested using
        automatically predicted tags from TnT.}
    \label{finalresults}
\end{table}
